---
title: "2023 NFL Losers Pool"
editor: source
description: Simulating NFL games for competing in a Losers Pool. 
categories:
  - R
  - Sports Analytics
  - Monte Carlo
format: 
  html: 
    code-fold: true
    code-tools: true
---


I am competing in another NFL Loser's Pool this year, and like past years 
I will be trying to use R to improve my chances in the pool. Feel free to 
read about the pool and my approach last year here: [NFL Losers Pool - 2022](https://www.jordanhutchings.com/blog/2022-09-07-nfl-losers-pool-2022/). 
This year, my goal is to improve and make it through week 4. 


## Generating Season Projections 

I've decided to look into creating my own ELO predictions. Specifically, 
I came across the `nflseedR` package, which appears to use ELO ratings to 
simulate the season - similar to the process done by [fivethirtyeight](https://fivethirtyeight.com/), 
which I have previosuly used. 

Let's first collect the NFL season data using `nflseedR`. 
We have the week, date, teams playing, and some of the betting 
markets data for each game. Here is a link to the 
[data dictionary](https://nflreadr.nflverse.com/articles/dictionary_schedules.html). 

```{r}
pacman::p_load(nflseedR, kableExtra, dplyr, ggplot2, RColorBrewer)
source(here::here("blog/2023-08-27-nfl-losers-pool-simulations/utils.R"))
knitr::opts_chunk$set(echo=TRUE, 
                      warning=FALSE, 
                      message=FALSE, 
                      fig.height=5, 
                      fig.width=8)

# setup params 
theme_set(
  theme_bw(10) + 
    theme(
      # panel.grid.minor = element_blank(), 
      axis.ticks.y = element_blank(),
      plot.title=element_text(size = 14, hjust = 0),
      plot.title.position = "plot",
      plot.subtitle=element_text(face="italic", size=12, margin=margin(b=12)),
      plot.caption=element_text(size=8, color="#7a7d7e", hjust = 0), 
      legend.position = "right", 
      legend.background = element_blank(),   
      legend.key = element_blank(), 
      strip.background = element_rect(fill = "#2C3E50"), 
      strip.text = element_text(color = "white")
    )
)

# read in projection data
sims = get_projections()

# format into data for methods
data = sims |> 
  mutate(team = case_when(away_percentage < home_percentage ~ away_team, 
                           away_percentage >= home_percentage ~ home_team), 
         pct = case_when(away_percentage < home_percentage ~ away_percentage, 
                         away_percentage >= home_percentage ~ home_percentage)) |> 
  select(week, team, pct) |> 
  filter(week %in% 1:12)
```


## Season win probabilities

```{r}
rbind(sims |> select(week, team = home_team, pct = home_percentage), 
      sims |> select(week, team = away_team, pct = away_percentage)) |> 
  mutate(team = reorder(team, pct), 
         text = sprintf("%.2f%%", pct)) |> 
  ggplot(aes(x = week, y = team, fill = pct)) + 
  geom_tile(color = "white") + 
  scale_fill_distiller(palette = "YlOrBr", labels = scales::percent_format()) + 
  scale_x_continuous(expand = c(0, 0), n.breaks = 18) + 
  scale_y_discrete(expand = c(0, 0)) + 
  # geom_text(aes(x = week, y = team, label = text), color = "white", size = 2) + 
  theme(panel.grid = element_blank(), 
        axis.title.y = element_blank()) + 
  labs(title = "Weekly Projections", 
       x = "Week Number", 
       fill = "Win Percent", 
       caption = "Notes:\nBlank tiles are bye weeks.\nPercentages simulated from nflseedR::simulate_nfl().")

```


## Pick sequencing

We are faced with the problem of when to pick each team. Given we simulated the season, 
and have win probabilities for each match-up, we can rely on some simple algorithms 
to compare possible pick orders. 

Consider for example the Greedy algorithm, and the Optimal algorithm. 
I am stretching the term "optimal" pretty thin here, however it is to refer to the 
set of picks which return the global minimum cumulative percent over a set of weeks. 

- Greedy algorithm: Pick the best team to win in each week, starting at the earliest week. 
This approach has no foresight and is all about trying to make it to see another day. 
- Optimal algorithm: Make a sequence of picks to minimize their cumulative win probabilities 
across all weeks. We are able to write some computer code and can use it to find the 
global minimum. 


```{r}
greedy = pick_greedy(data)
optimal = pick_optimal(data)

left_join(greedy, optimal, by = "week", suffix = c("_greedy", "_optimal")) |> 
  select(week, team_greedy, pct_greedy, team_optimal, pct_optimal) |> 
  kbl(caption = "Comparing Greedy and Optimal Algo", 
      col.names = c("Week", "Pick", "Win Prob.", "Pick", "Win Prob."),
      digits = 2) |> 
  add_header_above(c(" "=1, "Greedy Algorithm"=2, "Optimal Algo"=2)) |> 
  kable_paper("striped")

# plot differences
picks = rbind(greedy |> mutate(label = "Greedy"), 
              optimal |> mutate(label = "Optimal"))

avg_pct = sprintf("Average win percent: Greedy %.2f%% | Optimal %.2f%%", 
                  mean(greedy$pct) * 100, 
                  mean(optimal$pct) * 100)

# make plot
picks_cleaned <- picks %>%
  group_by(week) %>%
  mutate(
    d = lag(pct) - pct,
    y_pos = (lag(pct) + pct) / 2,
    y_label = ifelse(d == 0, "", paste0(round(d * 100, 1), "%")),
    team = ifelse(y_label == "" & !is.na(y_label), "", team)
  )

ggplot(picks_cleaned, aes(x = week, y = pct, color = label, label = y_label)) +
  geom_line(aes(group = week), color = "#e3e2e1", size = 2, alpha = 0.7) +
  geom_point(size = 3) +
  geom_text(aes(x = week, y = y_pos), color = "#414a4c", nudge_x = 0.5) +
  # ggrepel::geom_text_repel(
  #   aes(x = week, y = pct, label = team),
  #   nudge_x = 0.3, color = "#414a4c", direction = "y"
  # ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(n.breaks = 12) +
  scale_y_continuous(labels = scales::percent_format(), n.breaks = 7) +
  labs(
    title = "Week-by-Week Win Probabilities across Models",
    subtitle = avg_pct,
    caption = "Differences are interpreted as the additional risk taken on by selecting the OC model.\nData is updated weekly, and so predictions in plot are subject to change.",
    x = "Week Number",
    y = "Win Probability",
    color = "Model"
  ) +
  theme(
    legend.position = "bottom",
    legend.background = element_blank()
  )

```

In order to get to the global minimum, you weakly deviate from the greedy strategy. 
This therefore guarantees that the greedy strategy will weakly outperform the optimal 
strategy out of the gate. 

As we see with our projections, it isn't until week 11 that the optimal algorithm 
catches up. 

```{r}
picks |> 
  ggplot(aes(x = week, y = prob, color = label)) + 
  geom_point() + 
  geom_line(alpha = 0.7) + 
  scale_y_continuous(labels = scales::percent_format()) + 
  scale_x_continuous(n.breaks = 9, minor_breaks = NULL) + 
  scale_color_brewer(palette = "Dark2") + 
  labs(title = "Probability of reaching a given week by model", 
       x = "Week Number") + 
  theme(axis.title.y = element_blank(), 
        legend.position = c(0.9, 0.85), 
        legend.title = element_blank(), 
        legend.background = element_rect(color = "black"))
```


If the pool was about maximizing the likelihood of reaching the latest week in the pool, 
then the Optimal algorithm would be the overall best solution. The goal of the pool 
is however not to reach the latest week, but to outlast the other participants in 
the game. Therefore, there is room to improve on both these algorithms by using 
some game theory and psychology into how we can expect others to make their picks.


## Sanity Checks 

There should be a strong correlation with our model predictions and the betting markets. 
Lets check how those interact just to make sure our projections arent completely void. 

```{r}
betting_data = load_sharpe_games() |> 
  filter(season == "2023")

# is the underdog the same from projections and betting markets? Lets use the spread 
spread = betting_data |> 
  group_by(week) |> 
  summarise(underdog = case_when(spread_line < 0 ~ home_team, 
                             TRUE ~ away_team), 
            spread = abs(spread_line), 
            moneyline = pmin(home_moneyline, away_moneyline))

left_join(data, spread, by = c("team"="underdog", "week")) |> 
  arrange(week, moneyline)
  
  
  
```


## Predicting the picks of others

If we begin to consider team ownership, and the picking heuristics made by those in the pool, 
we can shift our objective from minimizing the risk of being eliminated to maximizing the 
likelihood of winning the pool. 

By predicting situations where there is a clear remaining pick for a substantial 
proportion of the pool, it can be advantageous to deviate from the crowd. These 
scenarios are where an upset could lead to a mass elimination, and the increase in 
the likelihood of winning offsets the increase in being eliminated that round. 

To emphisize this point, consider the following example: 

Suppose there are $N-1$ other players remaining in the pool, not counting yourself. 
Let's further assume all the other players will be picking the same game, say the worst team 
in the league playing the best team. 
To simplify things further, the pool ends after this game and the prize pool $\$M$ is 
to be split with the surviving players. Let's work out what the probability would 
have to be for it to be advantageous to pick the underdog. 

\begin{align*}
EV(\text{underdog}) &> E(\text{favourite}) \\
M \times \frac{1}{N} \times [\text{Prob. Favourite Wins}] &> M \times [\text{Prob. Underdog Wins}] \\ 
\frac{M\times p}{N} &> M\times (1-p) \\
p &> \frac{N}{N+1}
\end{align*}

Showing we should only follow the crowd if the probability of the team losing 
is very high. In the case where there are 7 players remaining in the pool, 
we would pick the favourite to win as long as the underdog team has a probability 
of winning of 12.5% or lower! 

We would not be in this extreme of a situation, however it still highlights 
the principal of deviating away from a team more likely to lose to increase the 
odds of winning the pool. 

### Common Heuristics

There are a handful of possible heuristics that others can use, and so it will 
be difficult to predict perfectly which is being used. 

1. Picking based on the spread
2. Picking based on who your faviourite team is playing
3. Picking based on fivethirtyeight or some other projections
4. Picking based on moneyline and other vegas odds

The process of predicting will get easier as the pool moves forward, as more 
players will have less teams remaining to pick from. 

